---
title: "Characterization of cytokine-mediated crosstalk within CTC-neutrophil clusters"
params:
  date: '`r format(Sys.Date(), "%B %d, %Y")`'
  output: pdf
  root_dir: ../../
  res_dir: results/cytokines_analysis_files
output:
  html_document:
    number_sections: yes
    theme: yeti
    toc: yes
    toc_depth: 4
    toc_float: yes
  pdf_document:
    number_sections: yes
    toc: yes
    toc_depth: '4'
  word_document:
    toc: yes
    toc_depth: '4'
---

# Summary

This report shows the co-expression analyses between cytokine and its receptors in CTC-WBC clusters. The cytokines gene sets were obtained from KEGG pathway named [Cytokine-cytokine receptor interaction](http://www.genome.jp/kegg-bin/show_pathway?hsa04060). 
  
# Setup

<!-- Setup Font for PDF -->
\fontfamily{phv}
\fontsize{9}{10}
\selectfont
\newpage


<!-- Analysis directories and global chunk options-->
```{r setup, include = FALSE}

# knitr options
options(knitr.duplicate.label = 'allow')


# Working directory
knitr::opts_knit$set(root.dir = normalizePath(params$root_dir))
setwd(params$root_dir)

# Other directories
if(!file.exists(params$res_dir))
  dir.create(params$res_dir, recursive = TRUE)
if(!file.exists('data/kegg'))
  dir.create('data/kegg', recursive = TRUE)

# Global chunk options
knitr::opts_chunk$set(fig.align='center', results='asis', echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, include = TRUE)

```

<!-- General Libraries -->
```{r libraries, include=FALSE}
# Src CRAN
library(tidyverse)
library(gridExtra)
library(grid)
library(circlize)
library(cowplot)
library(ggrepel)
library(ggsci)
library(wesanderson)
library(RColorBrewer)
library(pander)
library(DT)
library(parallel)
library(ComplexHeatmap)

# Src Bioconductor
library(scater)

```

<!-- PDF panderOptions for pandoc -->
```{r panderOptions, include=FALSE}

pander::panderOptions('table.alignment.default', 'right')
pander::panderOptions('table.alignment.rownames', 'left')
pander::panderOptions('table.split.table', Inf)
pander::panderOptions('table.caption.prefix', 'Table : ')

```

# Load data
<!-- Load featureCounts SCE -->
```{r}

sce.hs <- readRDS('data/sce_hs.rds')
colData(sce.hs) <- colData(sce.hs) %>% data.frame %>% 
  rownames_to_column('rnames') %>% 
  dplyr::mutate(
    donor = gsub("_.*", "", sample)
  ) %>% 
  column_to_rownames('rnames') %>% 
  DataFrame

sce.mm <- readRDS('data/sce_mm.rds')
colData(sce.mm) <- colData(sce.mm) %>% data.frame %>% 
  rownames_to_column('rnames') %>% 
  dplyr::mutate(
    donor = gsub("_.*", "", sample)
  ) %>% 
  column_to_rownames('rnames') %>% 
  DataFrame

```

<!-- Annotate Human SCE with RCA clusters -->
```{r}
sce <- sce.hs

# Get cluster ID from data
sce.clusterID <- colData(sce) %>% data.frame %>% 
  dplyr::mutate(clusterID = ifelse(clusterID == 'NA', NA, clusterID)) %>%
  collect %>% .[['clusterID']] %>% na.omit

# Get WBC-cluster from rca analysis
rca <- readRDS('results/rca/rca_hsa.rds')
rca.cellClust <- rca$cellClust

# Obtain CTC-cluster ID for rca analysis
rca.cID <- colData(sce)[names(rca.cellClust),] %>% data.frame %>% 
  dplyr::mutate(clusterID = ifelse(clusterID == 'NA', NA, clusterID)) %>%
  collect %>% .[['clusterID']]
rca.cellClust.Id <- rca.cellClust[!is.na(rca.cID)] # Subst only values with an assigned CTC-cluster ID
names(rca.cellClust.Id) <- na.omit(rca.cID)
rca.cellClust.Id <- cbind(clusterID = names(rca.cellClust.Id), rca.cluster.cID = as.character(rca.cellClust.Id)) %>%
  data.frame(stringsAsFactors = F) %>% 
  unique()

# Collapse cluster ID for duplicated names
use.df <- rca.cellClust.Id
use.df.res <- sapply(rca.cellClust.Id$clusterID %>% unique, function(id) {
  x.id <- use.df %>% dplyr::filter(clusterID == id)
  c.n <- x.id$rca.cluster.cID %>% unique %>% sort %>% paste(., collapse = ",")
  return(c(id, c.n))
  }) %>% 
  t() %>% 
  data.frame(stringsAsFactors = F) %>% 
  dplyr::rename(clusterID = X1, rca.cluster.cID = X2)
rca.cellClust.Id <- use.df.res

# Add to pheno data using CTC-cluster ID
p.useSce <- colData(sce) %>% data.frame %>% 
  rownames_to_column('keeprow') %>%
  left_join(rca.cellClust.Id, by = 'clusterID') %>%
  dplyr::mutate(rca.cluster = "No-eval") %>% 
  dplyr::mutate(rca.cluster = ifelse(rca.cluster=="No-eval" & !is.na(rca.cluster.cID), rca.cluster.cID, rca.cluster)) %>%
  dplyr::select(-rca.cluster.cID) %>%
  column_to_rownames('keeprow')

# add RCA using the sample name (original samples from RCA, some of them doesn't have an assigned CTC-cluster ID)
rca.cellClust <- rca.cellClust[names(rca.cellClust) %in% rownames(colData(sce))]

p.useSce[names(rca.cellClust), 'rca.cluster'] <- as.character(rca.cellClust[names(rca.cellClust)])

colData(sce) <- p.useSce %>% DataFrame

sce.hs <-sce
```

<!-- Annotate Mouse SCE with RCA clusters -->
```{r}
sce <- sce.mm

# Get cluster ID from data
sce.clusterID <- colData(sce) %>% data.frame %>% 
  dplyr::mutate(clusterID = ifelse(clusterID == 'NA', NA, clusterID)) %>%
  collect %>% .[['clusterID']] %>% na.omit

# Get WBC-cluster from rca analysis
rca <- readRDS('results/rca/rca_mmu.rds')
rca.cellClust <- rca$cellClust

# Obtain CTC-cluster ID for rca analysis
rca.cID <- colData(sce)[names(rca.cellClust),] %>%  data.frame %>% 
  dplyr::mutate(clusterID = ifelse(clusterID == 'NA', NA, clusterID)) %>%
  collect %>% .[['clusterID']]
rca.cellClust.Id <- rca.cellClust[!is.na(rca.cID)] # Subst only values with an assigned CTC-cluster ID
names(rca.cellClust.Id) <- na.omit(rca.cID)
rca.cellClust.Id <- cbind(clusterID = names(rca.cellClust.Id), rca.cluster.cID = as.character(rca.cellClust.Id)) %>%
  data.frame(stringsAsFactors = F) %>% 
  unique()

# Collapse cluster ID for duplicated names
use.df <- rca.cellClust.Id
use.df.res <-sapply(rca.cellClust.Id$clusterID %>% unique, function(id) {
  x.id <- use.df %>% dplyr::filter(clusterID == id)
  c.n <- x.id$rca.cluster.cID %>% unique %>% sort %>% paste(., collapse = ",")
  return(c(id, c.n))
  }) %>% 
  t() %>% 
  data.frame(stringsAsFactors = F) %>% 
  dplyr::rename(clusterID = X1, rca.cluster.cID = X2)
rca.cellClust.Id <- use.df.res

# Add to pheno data using CTC-cluster ID
p.useSce <- colData(sce) %>% data.frame %>% 
  rownames_to_column('keeprow') %>%
  left_join(rca.cellClust.Id, by = 'clusterID') %>%
  dplyr::mutate(rca.cluster = "No-eval") %>% 
  dplyr::mutate(rca.cluster = ifelse(rca.cluster=="No-eval" & !is.na(rca.cluster.cID), rca.cluster.cID, rca.cluster)) %>%
  dplyr::select(-rca.cluster.cID) %>%
  column_to_rownames('keeprow')

p.useSce[names(rca.cellClust), 'rca.cluster'] <- as.character(rca.cellClust[names(rca.cellClust)])


colData(sce) <- p.useSce %>% DataFrame

sce.mm <- sce

```




# Cytokines gene sets
* KEGG : Cytokine-cytokine receptor interaction
  + Human : http://www.genome.jp/kegg-bin/show_pathway?org_name=hsa&mapno=04060&mapscale=0.67&show_description=hide
  + Mouse : http://www.genome.jp/kegg-bin/show_pathway?org_name=mmu&mapno=04060&mapscale=0.67&show_description=show

<!-- Create data from KEGG : Cytokine-cytokine receptor interaction -->
```{r}

# KEGG : Cytokine-cytokine receptor interaction
# Human : http://www.genome.jp/kegg-bin/show_pathway?org_name=hsa&mapno=04060&mapscale=0.67&show_description=hide
# Mouse : http://www.genome.jp/kegg-bin/show_pathway?org_name=mmu&mapno=04060&mapscale=0.67&show_description=show
# I will use KEGGgraph (http://bioconductor.org/packages/release/bioc/html/KEGGgraph.html) to extract gene id and interactions 
# Pathways are stored and presented as graphs on the KEGG server side, where nodes are
# molecules (protein, compound, etc) and edges represent relation types between the nodes,
# e.g. activation or phosphorylation. 
# KEGGgraph requires the  KGML (KEGG XML) files that show the list of genes (nodes) and their interaction (edges)
kegg.data <- './data/kegg'

# HUMAN ---------------------------------------------------
res.file <- file.path(kegg.data, 'hsa04060.tsv.txt')
if(!file.exists(res.file)){
  library(KEGGgraph)
  library(KEGGREST)
  
  # 1. Get KGML file
  kgml.file <- file.path(kegg.data, 'hsa04060.kgml')
  mapkKGML <- retrieveKGML('hsa04060', organism="hsa", destfile=kgml.file, method="wget", quiet=TRUE)
  
  # 2. Parse to a graph
  mapkG <- parseKGML2Graph(mapkKGML, expandGenes=TRUE)
  mapkP <- parseKGML(mapkKGML)
  
  # 3. Convert nodes to gene names
  cl <- makeCluster(3, type="FORK")
  kegg.query <- parSapply(cl, nodes(mapkG), keggGet)
  stopCluster(cl)
  
  # 4. Extact gene names from the keeg query
  kegg.query.NAME <- lapply(kegg.query, function(x){
    strsplit(x$NAME, ", ") %>% unlist() %>% .[[1]]
  }) %>% unlist()
  
  # 5. Create data with Interactions
  mapkG.edges <- edges(mapkG)
  mat <- matrix(nrow = 0, ncol = 2, dimnames = list(c(), c('gene1', 'gene2')))
  for ( i in names(mapkG.edges)){
    if(length(mapkG.edges[[i]]) > 0)
      mat <- rbind(mat , cbind(i, mapkG.edges[[i]]))
  }
  mat[,1] <- kegg.query.NAME[mat[,1]]
  mat[,2] <- kegg.query.NAME[mat[,2]]
  
  data.frame(mat) %>% 
    write_tsv(path = res.file)
  
  detach("package:KEGGgraph", unload=TRUE) 
  detach("package:KEGGREST", unload=TRUE)
}


# Mouse ---------------------------------------------------
res.file <- file.path(kegg.data, 'mmu04060.tsv.txt')
if(!file.exists(res.file)){
  
  library(KEGGgraph)
  library(KEGGREST)
  
  # 1. Get KGML file
  kgml.file <- file.path(kegg.data, 'mmu04060.kgml')
  mapkKGML <- retrieveKGML('mmu04060', organism="mmu", destfile=kgml.file, method="wget", quiet=TRUE)
  
  # 2. Parse to a graph
  mapkG <- parseKGML2Graph(mapkKGML, expandGenes=TRUE)
  mapkP <- parseKGML(mapkKGML)
  
  # 3. Convert nodes to gene names
  cl <- makeCluster(6, type="FORK")
  kegg.query <- parSapply(cl, nodes(mapkG), keggGet)
  stopCluster(cl)
  
  # 4. Extact gene names from the keeg query
  kegg.query.NAME <- lapply(kegg.query, function(x){
    if(!is.null(x$NAME))
      strsplit(x$NAME, ", ") %>% unlist() %>% .[[1]]
  }) %>% unlist()
  
  # 5. Create data with Interactions
  mapkG.edges <- edges(mapkG)
  mat <- matrix(nrow = 0, ncol = 2, dimnames = list(c(), c('gene1', 'gene2')))
  for ( i in names(mapkG.edges)){
    if(length(mapkG.edges[[i]]) > 0)
      mat <- rbind(mat , cbind(i, mapkG.edges[[i]]))
  }
  mat[,1] <- kegg.query.NAME[mat[,1]]
  mat[,2] <- kegg.query.NAME[mat[,2]]
  
  data.frame(mat) %>% 
    write_tsv(path = res.file)
  
  detach("package:KEGGgraph", unload=TRUE) 
  detach("package:KEGGREST", unload=TRUE)
}


# Load the files-------------------------------------------
hs.cytokine.map <- read_tsv(file.path(kegg.data, 'hsa04060.tsv.txt')) %>% 
  dplyr::rename(cytokine = gene1, receptor = gene2) %>% 
  na.omit
mm.cytokine.map <- read_tsv(file.path(kegg.data, 'mmu04060.tsv.txt')) %>% 
  dplyr::rename(cytokine = gene1, receptor = gene2) %>% 
  na.omit
```


<!-- Generate orthologous pathway Mouse to Human -->
```{r}
kegg.data <- './data/kegg'

orth.file.mm <- file.path(kegg.data, 'mmu04060.orth.tsv.txt')
orth.file.hs <- file.path(kegg.data, 'hsa04060.orth.tsv.txt')

if(!file.exists(orth.file.mm) | !file.exists(orth.file.hs)){
  library(biomaRt)
  mouse.ensembl <- useMart("ensembl", dataset = "mmusculus_gene_ensembl")
  listAttributes(mouse.ensembl) %>% dplyr::filter(grepl('hsapiens', name))
  attributes <-  c("external_gene_name","hsapiens_homolog_associated_gene_name",
                   "hsapiens_homolog_orthology_confidence", "hsapiens_homolog_orthology_type", 
                   "hsapiens_homolog_perc_id", "hsapiens_homolog_perc_id_r1", 
                   "hsapiens_homolog_goc_score")
  
  orth.mm <-  getBM(attributes, filters="with_hsapiens_homolog", 
                    values=TRUE, mart = mouse.ensembl, uniqueRows=TRUE) 
  
  # Resolve ortholog_many2many and ortholog_one2many selecting by confidence, goc score and homology %
  orth.mm.unique <- orth.mm %>% 
    dplyr::filter(hsapiens_homolog_orthology_confidence == 1) %>% 
    dplyr::filter(hsapiens_homolog_goc_score >=90 ) %>%
    # dplyr::filter(hsapiens_homolog_associated_gene_name %in% hs.cytokine.map$cytokine) %>% 
    dplyr::arrange(desc(hsapiens_homolog_perc_id)) %>% 
    dplyr::select(external_gene_name, hsapiens_homolog_associated_gene_name) %>% 
    dplyr::filter(!duplicated(external_gene_name)) %>% 
    dplyr::filter(!duplicated(hsapiens_homolog_associated_gene_name))
  

  # Add orthology information to cytokine map  
  mm.cytokine.map.hs <- mm.cytokine.map %>% 
    # Add cytokines orthologous
    left_join(orth.mm.unique, by = c('cytokine'='external_gene_name')) %>% 
    dplyr::mutate(hsapiens_homolog_associated_gene_name = ifelse(is.na(hsapiens_homolog_associated_gene_name),
                                                                 toupper(cytokine),
                                                                 hsapiens_homolog_associated_gene_name)) %>% 
    dplyr::rename(cytokine.hs = hsapiens_homolog_associated_gene_name)%>% 
    # Add receptor orthologous
    left_join(orth.mm.unique, by = c('receptor'='external_gene_name')) %>% 
    dplyr::mutate(hsapiens_homolog_associated_gene_name = ifelse(is.na(hsapiens_homolog_associated_gene_name),
                                                                 toupper(receptor),
                                                                 hsapiens_homolog_associated_gene_name)) %>% 
    dplyr::rename(receptor.hs = hsapiens_homolog_associated_gene_name)
  
  
  # Select only shared interactions in human and mouse othologous
  mm.hs.cyr <- paste(mm.cytokine.map.hs$cytokine.hs, mm.cytokine.map.hs$receptor.hs, sep = '-')
  hs.cyr <- paste(hs.cytokine.map$cytokine, hs.cytokine.map$receptor, sep = '-')
  
  mm.cytokine.map.hs <- mm.cytokine.map.hs[mm.hs.cyr %in% hs.cyr,]
  hs.cytokine.map.mm <- hs.cytokine.map[hs.cyr %in% mm.hs.cyr,]
  
  # Write to a file
  mm.cytokine.map.hs %>% 
    write_tsv(path = orth.file.mm)
    
  hs.cytokine.map.mm %>% 
    write_tsv(path = orth.file.hs)

  detach("package:biomaRt", unload=TRUE) 
}

# Load the files-------------------------------------------
hs.cytokine.map.mm <- read_tsv(orth.file.hs)
mm.cytokine.map.hs <- read_tsv(orth.file.mm) 



```


# Prepare data
Analyses performed on detached CTC and **WBC-Neutrophils** from CTC-WBC clusters obtained from mouse models, CDX and patient samples

<!-- Mouse data -->
```{r}

# Select samples :
# WBC detached from a CTC  = 'CTC.wbc'
# CTC detached from a WBC  = 'CTC.cluster.WBC.ctc' | 'CTC.sinlge.WBC.ctc'
sce.chunk <- sce.mm
sample_cell_cat.sel <- c('CTC.wbc', 'CTC.cluster.WBC.ctc', 'CTC.sinlge.WBC.ctc')
sample.sel <- 
  colData(sce.chunk)  %>% data.frame %>% 
    dplyr::filter(
      organism == 'Mus musculus' 
      ) %>%
    dplyr::filter(
      sample_cell_cat %in% sample_cell_cat.sel
    ) %>% collect %>% .[['sample']]

sce.mouse <-   sce.chunk[,sample.sel]



# Annotate samples with both WBC and CTC detached
clusterID.paired <- 
  table(colData(sce.mouse)$clusterID ,
      colData(sce.mouse)$cell_type %>% as.character) %>% 
  as.data.frame.matrix %>%
  rownames_to_column('clusterID') %>%
  dplyr::filter(CTC > 0 & WBC >0) %>% 
  collect %>% .[['clusterID']]

is.paired <- colData(sce.mouse)$clusterID %in% clusterID.paired
colData(sce.mouse)[,'Paired_WBC_CTC'] <- is.paired

# Remove clusters samples that don't belong to the neutrophil cluster (clusters 0, 1 and 3 from RCA analysis)
clusterID.neutrophil <- colData(sce.mouse) %>% data.frame %>% dplyr::filter(rca.cluster == '1' | rca.cluster == '3' | rca.cluster == '0') %>% collect %>% .[['clusterID']]
use.clusters <- colData(sce.mouse)$clusterID %in% clusterID.neutrophil
sce.mouse <- sce.mouse[,use.clusters]

```

<!-- Patient data -->
```{r}

# Select samples :
# WBC detached from a CTC  = 'CTC.wbc'
# CTC detached from a WBC  = 'CTC.cluster.WBC.ctc' | 'CTC.single.WBC.ctc'
sce.chunk <- sce.hs
sample_cell_cat.sel <- c('CTC.wbc', 'CTC.cluster.WBC.ctc', 'CTC.single.WBC.ctc')
sample.sel <- 
  colData(sce.chunk)  %>% data.frame %>% 
    dplyr::filter(
      origin == 'patient' 
      ) %>%
    dplyr::filter(
      sample_cell_cat %in% sample_cell_cat.sel
    ) %>% collect %>% .[['sample']]

sce.patients.pure <-   sce.chunk[,sample.sel]


# Annotate samples with both WBC and CTC detached
clusterID.paired <- 
  table(colData(sce.patients.pure)$clusterID ,
      colData(sce.patients.pure)$cell_type %>% as.character) %>% 
  as.data.frame.matrix %>%
  rownames_to_column('clusterID') %>%
  dplyr::filter(CTC > 0 & WBC >0) %>% 
  collect %>% .[['clusterID']]

is.paired <- colData(sce.patients.pure)$clusterID %in% clusterID.paired
colData(sce.patients.pure)[,'Paired_WBC_CTC'] <- is.paired


# Remove clusters samples that don't belong to the neutrophil cluster (clusters 1 from RCA analysis)
clusterID.neutrophil <- colData(sce.patients.pure) %>% data.frame %>% dplyr::filter(rca.cluster == '1') %>% collect %>% .[['clusterID']]
use.clusters <- colData(sce.patients.pure)$clusterID %in% clusterID.neutrophil
sce.patients.pure <- sce.patients.pure[,use.clusters]


```

<!-- CDX data -->
```{r}

# Shared clusteID from human and mouse cells
shared.clusterID <- intersect(sce.hs$clusterID, sce.mm$clusterID)
shared.clusterID <- shared.clusterID[shared.clusterID != 'NA']

# From human, select samples :
# CTC detached from a WBC  = 'CTC.cluster.WBC.ctc' | 'CTC.single.WBC.ctc'
sce.chunk <- sce.hs
sample_cell_cat.sel <- c('CTC.cluster.WBC.ctc', 'CTC.single.WBC.ctc')
sample.sel <- 
  colData(sce.chunk)  %>% data.frame %>% 
    dplyr::filter(
      origin == 'xenograft' 
      ) %>%
    dplyr::filter(
      sample_cell_cat %in% sample_cell_cat.sel & 
      clusterID %in% shared.clusterID
    ) %>% 
  collect %>% .[['sample']]
sce.cdx.hs <-   sce.chunk[,sample.sel]

# From Mouse, select samples :
# WBC detached from a CTC  = 'CTC.wbc'
sce.chunk <- sce.mm
sample_cell_cat.sel <- c('CTC.wbc')
sample.sel <- 
  colData(sce.chunk)  %>% data.frame %>% 
    dplyr::filter(
      organism == 'Mus musculus' 
      ) %>%
    dplyr::filter(
      sample_cell_cat %in% sample_cell_cat.sel & 
      clusterID %in% shared.clusterID
    ) %>% collect %>% .[['sample']]
sce.cdx.mm <- sce.chunk[,sample.sel]


# Remove clusters samples that don't belong to the neutrophil cluster (clusters 0, 1 and 3 from RCA analysis)
clusterID.neutrophil <- colData(sce.cdx.mm) %>% data.frame %>% dplyr::filter(rca.cluster == '1' | rca.cluster == '3' | rca.cluster == '0') %>% collect %>% .[['clusterID']]
use.clusters.mm <- colData(sce.cdx.mm)$clusterID %in% clusterID.neutrophil
use.clusters.hs <- colData(sce.cdx.hs)$clusterID %in% clusterID.neutrophil
sce.cdx.mm <- sce.cdx.mm[,use.clusters.mm]
sce.cdx.hs <- sce.cdx.hs[,use.clusters.hs]
```

<!-- Create data with cytokine-receptor expression in mouse -->
```{r}

cytokine.map <- mm.cytokine.map.hs %>% data.frame
sce.chunk <- sce.mouse

# Remove non-paired samples
use.sample <- colData(sce.chunk)$Paired_WBC_CTC
sce.chunk <- sce.chunk[,use.sample]

# Select only Cytokine-Receptor pairs that are present in the data
use.cytokine <- cytokine.map$cytokine %in% rownames(sce.chunk)
use.receptor <- cytokine.map$receptor %in% rownames(sce.chunk)
cytokine.map <- cytokine.map[use.cytokine & use.receptor,]

# Sample annotation
sample.annot <- colData(sce.chunk)[,c('organism', 'donor', 'sample', 'clusterID', 'cell_type')] %>% data.frame %>% 
  dplyr::mutate(cell_type = as.character(cell_type))

# Correct clusterId with duplicated cell type (i.e: 1 WBC and 2 CTC samples), creating all combinations of CTC and WBC for the same clusterID
sample.annot.ctc <- sample.annot %>% dplyr::filter(cell_type == 'CTC')
res <- c()
for (ctc.sample in sample.annot.ctc$sample){
  organism <-  sample.annot[sample.annot$sample == ctc.sample, 'organism']
  donor <-  sample.annot[sample.annot$sample == ctc.sample, 'donor']
  clusterID <-  sample.annot[sample.annot$sample == ctc.sample, 'clusterID']
  samples.wbc <- sample.annot[sample.annot$clusterID == clusterID, ] %>% 
    dplyr::filter(cell_type == 'WBC') %>% 
    collect %>% .[['sample']]
  for(wbc.sample in samples.wbc){
    res <- rbind(res, 
                 c(organism = organism, donor = donor, CTC = ctc.sample, WBC = wbc.sample, clusterID = clusterID))
  }  
}
res <- res %>% data.frame(stringsAsFactors = FALSE)
res$clusterID <- make.names(res$clusterID, unique = T)

sample.annot.comb <- res

# Create data with cytokine and receptor expression
cyr.dat <- c()
i <- 1
for(i in 1:nrow(cytokine.map)){
  use.dat <- sample.annot.comb
    
  use.dat$cytokine <- cytokine.map[i, 'cytokine']
  use.dat$receptor <- cytokine.map[i, 'receptor']
  use.dat$cr.id <- paste(use.dat$cytokine, use.dat$receptor, sep = '-')

  use.dat$cytokine.wbc.expr <- unique(logcounts(sce.chunk)[use.dat$cytokine, use.dat$WBC])[1,]
  use.dat$cytokine.ctc.expr <- unique(logcounts(sce.chunk)[use.dat$cytokine, use.dat$CTC])[1,]
  use.dat$receptor.wbc.expr <- unique(logcounts(sce.chunk)[use.dat$receptor, use.dat$WBC])[1,]
  use.dat$receptor.ctc.expr <- unique(logcounts(sce.chunk)[use.dat$receptor, use.dat$CTC])[1,]

  cyr.dat <- rbind(cyr.dat, use.dat)
}

cyr.dat.mm <- cyr.dat

```

<!-- Create data with cytokine-receptor expression in human -->
```{r}

cytokine.map <- hs.cytokine.map.mm %>% data.frame
sce.chunk <- sce.patients.pure

# Remove non-paired samples
use.sample <- colData(sce.chunk)$Paired_WBC_CTC
sce.chunk <- sce.chunk[,use.sample]

# Select only Cytokine-Receptor pairs that are present in the data
use.cytokine <- cytokine.map$cytokine %in% rownames(sce.chunk)
use.receptor <- cytokine.map$receptor %in% rownames(sce.chunk)
cytokine.map <- cytokine.map[use.cytokine & use.receptor,]

# Sample annotation
sample.annot <- colData(sce.chunk)[,c('organism', 'donor', 'sample', 'clusterID', 'cell_type')] %>% data.frame %>% 
  dplyr::mutate(cell_type = as.character(cell_type))

# Correct clusterId with duplicated cell type (i.e: 1 WBC and 2 CTC samples), creating all combinations of CTC and WBC for the same clusterID
sample.annot.ctc <- sample.annot %>% dplyr::filter(cell_type == 'CTC')
res <- c()
for (ctc.sample in sample.annot.ctc$sample){
  organism <-  sample.annot[sample.annot$sample == ctc.sample, 'organism']
  donor <-  sample.annot[sample.annot$sample == ctc.sample, 'donor']
  clusterID <-  sample.annot[sample.annot$sample == ctc.sample, 'clusterID']
  samples.wbc <- sample.annot[sample.annot$clusterID == clusterID, ] %>% 
    dplyr::filter(cell_type == 'WBC') %>% 
    collect %>% .[['sample']]
  for(wbc.sample in samples.wbc){
    res <- rbind(res, 
                 c(organism = organism, donor = donor, CTC = ctc.sample, WBC = wbc.sample, clusterID = clusterID))
  }  
}
res <- res %>% data.frame(stringsAsFactors = FALSE)
res$clusterID <- make.names(res$clusterID, unique = T)

sample.annot.comb <- res

# Create data with cytokine and receptor expression
cyr.dat <- c()
i <- 1
for(i in 1:nrow(cytokine.map)){
  use.dat <- sample.annot.comb
    
  use.dat$cytokine <- cytokine.map[i, 'cytokine']
  use.dat$receptor <- cytokine.map[i, 'receptor']
  use.dat$cr.id <- paste(use.dat$cytokine, use.dat$receptor, sep = '-')

  use.dat$cytokine.wbc.expr <- unique(logcounts(sce.chunk)[use.dat$cytokine, use.dat$WBC])[1,]
  use.dat$cytokine.ctc.expr <- unique(logcounts(sce.chunk)[use.dat$cytokine, use.dat$CTC])[1,]
  use.dat$receptor.wbc.expr <- unique(logcounts(sce.chunk)[use.dat$receptor, use.dat$WBC])[1,]
  use.dat$receptor.ctc.expr <- unique(logcounts(sce.chunk)[use.dat$receptor, use.dat$CTC])[1,]

  cyr.dat <- rbind(cyr.dat, use.dat)
}

cyr.dat.hs <- cyr.dat

```

<!-- Create data with cytokine-receptor expression in CDX -->
```{r}

# samples.paired
cytokine.map <- mm.cytokine.map.hs %>% data.frame
sce.chunk.mm <- sce.cdx.mm
sce.chunk.hs <- sce.cdx.hs

# Remove non-paired samples
clusterID.paired <- intersect(colData(sce.chunk.hs)$clusterID, colData(sce.chunk.mm)$clusterID)
paired.samples.mm <- colData(sce.chunk.mm) %>% data.frame %>% 
  dplyr::filter(cell_type == 'WBC') %>% 
  dplyr::filter(clusterID %in% clusterID.paired) %>% 
  collect %>% .[['sample']]
paired.samples.hs <- colData(sce.chunk.hs) %>% data.frame %>% 
  dplyr::filter(cell_type == 'CTC') %>% 
  dplyr::filter(clusterID %in% clusterID.paired) %>% 
  collect %>% .[['sample']]
sce.chunk.mm <- sce.chunk.mm[,paired.samples.mm]
sce.chunk.hs <- sce.chunk.hs[,paired.samples.hs]

# Convert gene names in mouse to human
# rownames(sce.chunk.mm) <- toupper(rownames(sce.chunk.mm))

# Select only Cytokine-Receptor pairs that are present in the data
use.cytokine.mm <- cytokine.map$cytokine %in% rownames(sce.chunk.mm)
use.receptor.mm <- cytokine.map$receptor %in% rownames(sce.chunk.mm)
use.cytokine.hs <- cytokine.map$cytokine.hs %in% rownames(sce.chunk.hs)
use.receptor.hs <- cytokine.map$receptor.hs %in% rownames(sce.chunk.hs)
cytokine.map <- cytokine.map[use.cytokine.mm & use.cytokine.hs & use.receptor.mm & use.receptor.hs,]

# Sample annotation
sample.annot.hs <- colData(sce.chunk.hs)[,c('organism', 'donor', 'sample', 'clusterID', 'cell_type')] %>% data.frame %>% 
  dplyr::mutate(cell_type = as.character(cell_type))
sample.annot.mm <- colData(sce.chunk.mm)[,c('organism', 'donor', 'sample', 'clusterID', 'cell_type')] %>% data.frame %>% 
  dplyr::mutate(cell_type = as.character(cell_type))

# Correct clusterId with duplicated cell type (i.e: 1 WBC and 2 CTC samples), creating all combinations of CTC and WBC for the same clusterID
sample.annot.ctc <- sample.annot.hs %>% dplyr::filter(cell_type == 'CTC')
sample.annot.wbc <- sample.annot.mm %>% dplyr::filter(cell_type == 'WBC')
res <- c()
for (ctc.sample in sample.annot.ctc$sample){
  organism <- 'cdx'
  donor <-  sample.annot.ctc[sample.annot.ctc$sample == ctc.sample, 'donor']
  clusterID <-  sample.annot.ctc[sample.annot.ctc$sample == ctc.sample, 'clusterID']
  samples.wbc <- sample.annot.wbc[sample.annot.wbc$clusterID == clusterID, ] %>% 
    collect %>% .[['sample']]
  for(wbc.sample in samples.wbc){
    res <- rbind(res, 
                 c(organism = organism, donor = donor, CTC = ctc.sample, WBC = wbc.sample, clusterID = clusterID))
    }
}
res <- res %>% data.frame(stringsAsFactors = FALSE)
res$clusterID <- make.names(res$clusterID, unique = T)

sample.annot.comb <- res

# Create data with cytokine and receptor expression
cyr.dat <- c()
i <- 1
for(i in 1:nrow(cytokine.map)){
  use.dat <- sample.annot.comb
    
  use.cytokine.mm <- cytokine.map[i, 'cytokine']
  use.receptor.mm <- cytokine.map[i, 'receptor']
  use.cytokine.hs <- cytokine.map[i, 'cytokine.hs']
  use.receptor.hs <- cytokine.map[i, 'receptor.hs']
  
  use.dat$cytokine <- use.cytokine.hs
  use.dat$receptor <- use.receptor.hs
  use.dat$cr.id <- paste(use.dat$cytokine, use.dat$receptor, sep = '-')

  use.dat$cytokine.wbc.expr <- logcounts(sce.chunk.mm)[use.cytokine.mm, use.dat$WBC]
  use.dat$cytokine.ctc.expr <- logcounts(sce.chunk.hs)[use.cytokine.hs, use.dat$CTC]
  use.dat$receptor.wbc.expr <- logcounts(sce.chunk.mm)[use.receptor.mm, use.dat$WBC]
  use.dat$receptor.ctc.expr <- logcounts(sce.chunk.hs)[use.receptor.hs, use.dat$CTC]

  cyr.dat <- rbind(cyr.dat, use.dat)
}

cyr.dat.cdx <- cyr.dat

```

<!-- Merge mouse, human and cdx cytokine-receptor expression data-->
```{r}

cyr.dat.mm <- cyr.dat.mm %>% 
  left_join(mm.cytokine.map.hs) %>% 
  dplyr::mutate(
    cytokine = cytokine.hs,
    receptor = receptor.hs,
    cr.id = paste(cytokine, receptor, sep = '-')
    ) %>% 
  dplyr::select(-cytokine.hs, -receptor.hs)


cyr.dat <- rbind(cyr.dat.hs, cyr.dat.mm, cyr.dat.cdx)

```

# Figures
<!-- Configuration -->
```{r}

logCPM.thrs <- log2(5)

cyr.dycho <- cyr.dat %>% 
  dplyr::mutate(
    cytokine.wbc.expr = cytokine.wbc.expr > logCPM.thrs,
    cytokine.ctc.expr = cytokine.ctc.expr > logCPM.thrs,
    receptor.wbc.expr = receptor.wbc.expr > logCPM.thrs,
    receptor.ctc.expr = receptor.ctc.expr > logCPM.thrs
  )

```

## Extended Data Fig. 6a
Heatmap showing the transcriptional landscape of cytokines and corresponding receptors expressed in at least 20% of CTC-neutrophil clusters.
```{r, fig.height=5, fig.width=12}

cyr.use <- cyr.dycho %>% 
  dplyr::mutate(
    cyr.cat = 'None',
    cyr.cat = ifelse(cytokine.wbc.expr, 'Cytokine Expressed', cyr.cat),
    cyr.cat = ifelse(receptor.ctc.expr, 'Receptor Expressed', cyr.cat),
    cyr.cat = ifelse(cytokine.wbc.expr & receptor.ctc.expr, 'Cytokine-Receptor Expressed', cyr.cat)
  )

cyr.spread <- cyr.use %>% dplyr::select(clusterID, cr.id, cyr.cat) %>%  spread(key = clusterID, value = cyr.cat)
rownames(cyr.spread) <- cyr.spread$cr.id
cyr.spread$cr.id <- NULL
cyr.spread[is.na(cyr.spread)] <- 'None'
mat <- as.matrix(cyr.spread)

# Heatmap
# Rows sorted by : Number of samples with  'Cytokine-Receptor Expressed'
# Columns sorted by number of combinations scaled by the combination order
cyr.row <- rowSums(mat == 'Cytokine-Receptor Expressed')
row.ord <- order(cyr.row, decreasing = TRUE)
col.score <- apply(mat[row.ord,], 2, function(x){
    x = x %>% factor %>% plyr::revalue(c('Cytokine-Receptor Expressed' = 10, 'Cytokine Expressed'= 3, 'Receptor Expressed' = 3, 'None' = -1)) %>% as.character %>% as.numeric()
    sum(x*(rev(seq_along(x))^100))
})
col.names.ord <- sort(col.score, decreasing = T) %>% names()
heat.colors <- c('Cytokine-Receptor Expressed' = "#BC3C29FF", 'Cytokine Expressed'= "#f6e8c3", 'Receptor Expressed' = "#79AF97B2", 'None' = "grey90")


# Columns sorted by donor
col.names.ord <- cyr.use %>% dplyr::filter(cyr.use$clusterID %in% colnames(cyr.spread)) %>% dplyr::select(donor, clusterID) %>% unique %>% dplyr::arrange(donor) %>% collect %>% .[['clusterID']]


mat.sel <- mat[row.ord,col.names.ord]
use.rows <- (rowSums(mat.sel == 'Cytokine-Receptor Expressed') / ncol(mat.sel))>=0.20
mat.sel <- mat.sel[use.rows,]


# Col annotation
sampleAnnot <- cyr.dycho %>% dplyr::select(organism, donor, CTC, WBC, clusterID) %>% unique %>% 
  dplyr::mutate(
    organism.ord = organism,
    organism.ord = ifelse(organism.ord == 'Homo sapiens', '1.Homo sapiens', organism.ord),
    organism.ord = ifelse(organism.ord == 'cdx', '2.cdx', organism.ord),
    organism.ord = ifelse(organism.ord == 'Mus musculus', '3.Mus musculus', organism.ord),
    donor.org = paste0(donor, " (", organism, ")"),
    org.donor = paste0(organism.ord, ": ", donor)
    ) %>% 
  dplyr::arrange(org.donor)
rownames(sampleAnnot) <- sampleAnnot$clusterID
sampleAnnot <- sampleAnnot %>% dplyr::select(donor.org)
use.rows <- rownames(sampleAnnot) %in% colnames(mat.sel)
sampleAnnot <- sampleAnnot[use.rows,,drop = FALSE]
donor.levels <- unique(sampleAnnot$donor.org)
donor.col <- pal_uchicago(alpha = 0.7)(length(donor.levels))
names(donor.col) <- donor.levels
ha.col <-  HeatmapAnnotation(
  sampleAnnot, 
  col = list(
    donor.org = donor.col
  ),
  annotation_legend_param = list(
    donor.org = list( 
      labels_gp = gpar(fontsize = 13),
      # legend_direction = "horizontal",
      grid_height = unit(8, "mm"),
      grid_width = unit(8, "mm"),
      title = "",
      nrow = 1
      )
  )
)


# Core heatmap
mat.sel <- mat.sel[,rownames(sampleAnnot)]
ht <- Heatmap(
  mat.sel,
  row_names_side = "left", 
  col = heat.colors,
  rect_gp = gpar(col = 'grey70'),
  name = '',
  top_annotation = ha.col,
  show_column_names = FALSE,
  heatmap_legend_param = list(
    labels_gp = gpar(fontsize = 13),
    legend_direction = "horizontal",
    grid_height = unit(8, "mm"),
    grid_width = unit(8, "mm"),
    title_gp = gpar(fontsize = 13, fontface = 'bold'),
    nrow = 2
  )
)

# draw(ht, heatmap_legend_side = "bottom", annotation_legend_side = "top")

# Row Annotations
n.cyr.row <- 100*rowSums(mat.sel == "Cytokine-Receptor Expressed") / ncol(mat.sel)
n.cyr.row.col <- rep("grey90", length(n.cyr.row))

ha.row <-  rowAnnotation(
  n.samples = row_anno_barplot(n.cyr.row, axis = TRUE, gp = gpar(fill = "grey90")),
  annotation_width = unit(2, "cm"))


draw(ht + ha.row, heatmap_legend_side = "bottom", annotation_legend_side = "top")

decorate_annotation("n.samples", {
    grid.text("Percent of pairs", unit(10, "mm"), unit(-14, "mm"), just = "bottom", rot = 0, gp = gpar(cex = 0.5))
})

```

## Extended Data Fig. 6b
Heatmap showing the transcriptional landscape of cytokines receptors and corresponding cytokines expressed in at least 20% of CTC-neutrophil clusters.
```{r, fig.height=5.5, fig.width=12}

cyr.use <- cyr.dycho %>% 
  dplyr::mutate(
    cyr.cat = 'None',
    cyr.cat = ifelse(cytokine.ctc.expr, 'Cytokine Expressed', cyr.cat),
    cyr.cat = ifelse(receptor.wbc.expr, 'Receptor Expressed', cyr.cat),
    cyr.cat = ifelse(cytokine.ctc.expr & receptor.wbc.expr, 'Cytokine-Receptor Expressed', cyr.cat)
  )

cyr.spread <- cyr.use %>% dplyr::select(clusterID, cr.id, cyr.cat) %>%  spread(key = clusterID, value = cyr.cat)
rownames(cyr.spread) <- cyr.spread$cr.id
cyr.spread$cr.id <- NULL
cyr.spread[is.na(cyr.spread)] <- 'None'
mat <- as.matrix(cyr.spread)


# Heatmap
# Rows sorted by : Number of samples with  'Cytokine-Receptor Expressed'
# Columns sorted by number of combinations scaled by the combination order
cyr.row <- rowSums(mat == 'Cytokine-Receptor Expressed')
row.ord <- order(cyr.row, decreasing = TRUE)
col.score <- apply(mat[row.ord,], 2, function(x){
    x = x %>% factor %>% plyr::revalue(c('Cytokine-Receptor Expressed' = 10, 'Cytokine Expressed'= 3, 'Receptor Expressed' = 3, 'None' = -1)) %>% as.character %>% as.numeric()
    sum(x*(rev(seq_along(x))^100))
})
col.names.ord <- sort(col.score, decreasing = T) %>% names()
heat.colors <- c('Cytokine-Receptor Expressed' = "#BC3C29FF", 'Cytokine Expressed'= "#f6e8c3", 'Receptor Expressed' = "#79AF97B2", 'None' = "grey90")


# Columns sorted by donor
col.names.ord <- cyr.use %>% dplyr::filter(cyr.use$clusterID %in% colnames(cyr.spread)) %>% dplyr::select(donor, clusterID) %>% unique %>% dplyr::arrange(donor) %>% collect %>% .[['clusterID']]


mat.sel <- mat[row.ord,col.names.ord]
use.rows <- (rowSums(mat.sel == 'Cytokine-Receptor Expressed') / ncol(mat.sel))>=0.20
mat.sel <- mat.sel[use.rows,]


# Col annotation
sampleAnnot <- cyr.dycho %>% dplyr::select(organism, donor, CTC, WBC, clusterID) %>% unique %>% 
  dplyr::mutate(
    organism.ord = organism,
    organism.ord = ifelse(organism.ord == 'Homo sapiens', '1.Homo sapiens', organism.ord),
    organism.ord = ifelse(organism.ord == 'cdx', '2.cdx', organism.ord),
    organism.ord = ifelse(organism.ord == 'Mus musculus', '3.Mus musculus', organism.ord),
    donor.org = paste0(donor, " (", organism, ")"),
    org.donor = paste0(organism.ord, ": ", donor)
    ) %>% 
  dplyr::arrange(org.donor)
rownames(sampleAnnot) <- sampleAnnot$clusterID
sampleAnnot <- sampleAnnot %>% dplyr::select(donor.org)
use.rows <- rownames(sampleAnnot) %in% colnames(mat.sel)
sampleAnnot <- sampleAnnot[use.rows,,drop = FALSE]
donor.levels <- unique(sampleAnnot$donor.org)
donor.col <- pal_uchicago(alpha = 0.7)(length(donor.levels))
names(donor.col) <- donor.levels
ha.col <-  HeatmapAnnotation(
  sampleAnnot, 
  col = list(
    donor.org = donor.col
  ),
  annotation_legend_param = list(
    donor.org = list( 
      labels_gp = gpar(fontsize = 13),
      # legend_direction = "horizontal",
      grid_height = unit(8, "mm"),
      grid_width = unit(8, "mm"),
      title = "",
      nrow = 1
      )
  )
)


# Core heatmap
mat.sel <- mat.sel[,rownames(sampleAnnot)]
ht <- Heatmap(
  mat.sel,
  row_names_side = "left", 
  col = heat.colors,
  rect_gp = gpar(col = 'grey70'),
  name = '',
  top_annotation = ha.col,
  show_column_names = FALSE,
  heatmap_legend_param = list(
    labels_gp = gpar(fontsize = 13),
    legend_direction = "horizontal",
    grid_height = unit(8, "mm"),
    grid_width = unit(8, "mm"),
    title_gp = gpar(fontsize = 13, fontface = 'bold'),
    nrow = 2
  )
)


# Row Annotations
n.cyr.row <- 100*rowSums(mat.sel == "Cytokine-Receptor Expressed") / ncol(mat.sel)
n.cyr.row.col <- rep("grey90", length(n.cyr.row))

ha.row <-  rowAnnotation(
  n.samples = row_anno_barplot(n.cyr.row, axis = TRUE, gp = gpar(fill = "grey90")),
  annotation_width = unit(c(2, 2), "cm"))


draw(ht + ha.row, heatmap_legend_side = "bottom", annotation_legend_side = "top")

decorate_annotation("n.samples", {
    grid.text("Percent of pairs", unit(10, "mm"), unit(-14, "mm"), just = "bottom", rot = 0, gp = gpar(cex = 0.5))
})

```

